---
title: "Disappearance of data related to the 'Mysterious spike' of deaths among people under 45 years of age in New Brunswick following vaccination"
subtitle: "Investigation report"
output:
  pdf_document: default
  self_contained: yes
  mode: selfcontained
  always_allow_html: true
  html_document: default
  # fig_caption: yes
  # toc: no
  # toc_depth: 2
  # toc_float: no
  # number_sections: false
# date: 'www.IVIM.ca/data/reports'
date: 2022-05-29
# author: Prepared by [IVIM](https://IVIM.ca) on 2022-05-29
# author: [Independently Verified Information Machine](https://IVIM.ca) 
author: Dr. Dmitry Gorodnichy
---




## Observation
<!-- ## Summary -->

The data published by the Statistics Canada  in its April's release of "Death counts, by age and sex" dataset  showed   significant   (by a factor of 2 to 4) increase of deaths  in New Brunswick in  age group 0 to 44 after June 2021 - i.e. one month after   the vaccination has been extended to this age group of the province.


In its next month's release of the dataset, Statistics Canada has removed the portion of data that showed this increase.

This Investigation Report provides details on what exactly has been reported in April and  what has been removed in May 

<!-- and hows how to recover the deleted portion data from other the remain data portions -->



<!-- : from  5 deaths per week, observed prior to June, which is equal to historical three year average, to over 20 deaths per week in November   -->



<!-- In its release of 2022-05-12 the data for  'all sexes' in age group 0 to 44 after 6 June 2021 have been erased, i.e.. original values that were published on 2022-04-14 have been replaced with blanks ('NA' (Not available). -->
<!-- This is exactly when a peak in deaths in this population of new-brunswicker was observed and commented on by the province health minister. -->

### Additional references and analysis

<!-- ## Background -->

<!-- See the following articles for more details about the observation: -->
<!-- and YouTube recording for the background: -->

- CBC article: "Mysterious spike in N.B. deaths in 2021 not due to COVID, health minister says", Posted: May 13, 2022, URL: https://www.cbc.ca/amp/1.6451680

- «"Mysterious spike" in excess deaths in New Brunswick: Now you see it. Now you don't»,   May 21 <https://ivim.substack.com/p/mysterious-spike-in-excess-deaths>
- «"Killer on the road" in New Brunswick: killing mainly under 44. CBC started talking about it. StatsCan removed data », 
May 19:  <https://ivim.substack.com/p/killer-on-the-road-in-new-brunswick>

<!-- - Recording of the "Open Canada Data" seminars on 2022-05-20 : <https://youtu.be/Q3J_P5yk_kM> -->

## Data source 

Source: Statistics Canada - Table 13-10-0768-01. Provisional weekly death counts, by age group and sex  (Updated monthly)

- URL: <https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310076801>  
- Data Releases: 2022-04-14, 2022-05-12    
- Cached data: <https://github.com/IVI-M/std-data/tree/main/ca/statcan-deaths-by-age>





```{r -globals, include=F}
knitr::opts_chunk$set(echo = F, message=FALSE, warning=FALSE)
source("00_common.R")


# > Read Deaths By age ----

# https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310076801

if (F) {
  dt <- cansim::get_cansim("13-10-0768-01") %>%  setDT() 
  
    fwrite(dt, "data/13100768-2022-06-09.csv")
  # fwrite(dt, "data/13100768-since2018-2022-05-12.csv")
  saveRDS(dt, paste0("3-10-0768-01-cached_on-2022-May-19.Rds")) # save locally as compressed Rds file
  dt <- dt[, c("Date", "GEO", "val_norm","Age at time of death", "Sex", "Characteristics")] 
}
if (F) {
  dt <- fread("data/13100768-2022-04-14.csv")
  dt <- fread("data/13100768-2022-05-12.csv")
  dt %>% names
  dt <- dt[, c("REF_DATE", "GEO", "VALUE","Age at time of death", "Sex")] 
  setnames(dt, "VALUE", "value")
  setnames(dt, "REF_DATE", "Date")
  setnames(dt, "Age at time of death", "Age")
  dt[, GEO := gsub(", place of occurrence", "", GEO)]
  dt[, Age := gsub("Age at time of death, ", "", Age)]
  dt <- dt[Date>=ymd("2018-01-01")]
  # dt <- dt[Sex=="Both sexes"]
  
  fwrite(dt, "data13100768-since2018-2022-05-12-all-sexes.csv")
  dt <- dt[GEO=="New Brunswick"]
  fwrite(dt, "13100768-NB-since2018-2022-04-14-all-sexes.csv")
  fwrite(dt, "13100768-NB-since2018-2022-05-12-all-sexes.csv")
  fwrite(dt, "13100768-NB-since2018-2022-06-09-all-sexes.csv")
}

dt1 <- fread("13100768-NB-since2018-2022-04-14-all-sexes.csv")
dt <- fread("13100768-NB-since2018-2022-05-12-all-sexes.csv")

# dtHistorical <- dt [Date >= ymd("2018-01-01") & Date <= ymd("2019-12-31") ][ , week:=week(Date)
#   ][ ,.(historical=as.integer(mean(value, na.rm=T))), by=.(GEO, Age, week, Sex) ]
# dtHistorical[, historical.ave := frollmean(historical, 8, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ]
# # dtHistorical$Date <- lubridate::ymd( "2014-01-01" ) + lubridate::weeks( df$Week - 1 )

# dtHistorical %>% ggplot()+geom_line(aes(Date, historical.ave), alpha=0.3)

releasedDeaths <- function(dt, release="2022-05-12", weeks_ave=6) {
  
  dtHistorical <- dt [Date >= ymd("2018-01-01") & Date <= ymd("2019-12-31") ][ , week:=week(Date)
  ][ ,.(historical=as.integer(mean(value, na.rm=T))), by=.(GEO, Age, week, Sex) ]
  
  dtPandemic <- dt [Date >= ymd("2020-01-01") ][, week:=week(Date)]
  # dtPandemic[, deaths := frollmean(value, weeks_ave, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ]
  # dtPandemic$release=release 
  
  dtExcess <- dtHistorical[dtPandemic, on=c( "GEO" , "week",  "Age", "Sex" ) ]

  dtExcess[, deaths := frollmean(value, 6, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ]
  dtExcess[, historical.ave := frollmean(historical, 8, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ]
  dtExcess$release=release
  dtExcess
}

plotDeaths <- function(dtExcess, release="2022-05-12") {
  
  g <- ggplot(dtExcess[Sex=="Both sexes"]) + 
    geom_vline(xintercept = ymd("2020-12-14"), linetype=1, alpha=0.4) +
    geom_vline(xintercept = ymd("2021-05-01"), linetype=1, size=2, alpha=0.4) +
    # geom_line(aes(Date, deathsMonthly/30*7), data=dt1[Date> "2020-01-01"], col="red",  alpha=0.8) +
    geom_line(aes(Date, value), col="red", alpha=0.3) +
    geom_line(aes(Date, deaths), col="red") +
    geom_line(aes(Date, historical.ave), col="black", alpha=0.5) +
    
    
    facet_grid(Age~release, scales = "free") +
    labs( title = "Mortality rate before and after vaccination", 
          subtitle=paste0("Region: New Brunswick. Data Release: ", release),
          
          x = "Black line: pre-pandemic 2018-2019 average. Thick line: six-week average",
          # y = ifelse(F,  "Deaths per million per week", "Deaths per week"),
          y="Deaths per week", # 13-10-0768-01 (age); 13-10-0810-01 (cause)
          caption = "Source: Statistics Canada - Table 13-10-0768-01\n Generated by Open Canada Deaths Tracker App (ivim.ca/app)"     )  
  
  # ggsave(paste0("NB-excess-deaths_asof_2022-05-12.png"), width = 12, height = 9)

  g
  
}


```



## Visual summary


<!-- ## Comparative graphs -->


<!-- ### Plots for printing -->

Vertical lines indicate the date when vaccination started with high-risk population (December 14, 2020) and (thick line) when vaccination was extended to general population  (May 1, 2021)


```{r out.width='100%', out.height='66%'}

release="2022-04-14"
dtExcess <- releasedDeaths (dt1, release, weeks_ave=6) 
# g <- plotDeaths(dtExcess, release="2022-04-14") 
# g

release="2022-05-12"
dtExcess1 <- releasedDeaths (dt, release, weeks_ave=6) 
# g1 <- plotDeaths(dtExcess1, release="2022-05-12") 
# g1

dtDiff <- rbind (dtExcess, dtExcess1)
plotDeaths(dtDiff, release="2022-05-12 vs. 2022-04-14") 

```


<!-- ### Interactive plots -->

<!-- Hover over the image for details and selection -->

<!-- ```{r  include=FALSE} -->
<!-- g1 %>% ggplotly -->
<!-- g %>% ggplotly -->
<!-- ``` -->


<!-- ```{r include=FALSE} -->
<!-- dt1 %>% datatable.title ("Total Deaths in New Brunswick before and after vaccination. Data release: 2022-04-14") -->
<!-- ``` -->



<!-- #### Open questions -->
<!-- #### Next steps -->
<!-- ##  Revering removed data from other data -->
<!-- ## Related consideration -->

## Recommendations

The dataset has several fields (columns) that are related to each other. In particular:

- 'All ages' values should equal to the sum of deaths in all age groups. In April release of the data, this appears to be the case. In May release it is not.

<!-- - 'Both sexes' values should be equal to the sum of deaths in both sexes. This does not appear to be the case. It is not clear how or why this could have been possible.  -->

Therefore, assuming the integrity of the other (unremoved) parts of dataset, it should possible to recover the deleted portion data for age group  0 to 44 by simply subtracting data for all other groups data from "All ages" data.

<!-- Hence further investigation and monitoring of these data is recommended -->



## Appendix: Raw Data preview



<!-- ## Data preview -->

Extracts below show :

- the first and last five lines of the dataset after 2018-01-01, 
- the date of the last entry, 
- the date of the last numeric entry for each age group (i.e. which is when entry is a  number, not NA - Not Available)



\newpage

### Data Release: 2022-04-14


```{r print dt}
dt1 

# Release date: 2022-05-12
# Loaded On "2022-05-19"
dt1 [, max(Date, na.rm = T), by=GEO] #2022-03-05
dt1 [!is.na(value) & GEO=="New Brunswick", max(Date, na.rm = T), by=.(Age,Sex)]
# 4:     0 to 44 years Both sexes 2021-06-05
# 5:     0 to 44 years      Males 2021-10-09
# 6:     0 to 44 years    Females 2021-11-13
# 7:    45 to 64 years Both sexes 2021-11-20

# Release date: 2022-04-14 . 2022-02-05
# 4:     0 to 44 years Both sexes 2021-10-23
# 5:     0 to 44 years      Males 2021-10-23
# 6:     0 to 44 years    Females 2021-10-23
# 7:    45 to 64 years Both sexes 2021-10-23
```


\newpage

### Data Release: 2022-05-12


```{r may, echo=FALSE, message=FALSE, warning=FALSE}


dt 
# Release date: 2022-05-12
# Loaded On "2022-05-19"
dt [, max(Date, na.rm = T), by=GEO] #2022-03-05
dt [!is.na(value) & GEO=="New Brunswick", max(Date, na.rm = T), by=.(Age,Sex)]
# 4:     0 to 44 years Both sexes 2021-06-05
# 5:     0 to 44 years      Males 2021-10-09
# 6:     0 to 44 years    Females 2021-11-13
# 7:    45 to 64 years Both sexes 2021-11-20

# Release date: 2022-04-14 . 2022-02-05
# 4:     0 to 44 years Both sexes 2021-10-23
# 5:     0 to 44 years      Males 2021-10-23
# 6:     0 to 44 years    Females 2021-10-23
# 7:    45 to 64 years Both sexes 2021-10-23

```


