---
# title: "New Brunswick excess deaths before and after vaccination"
# subtitle: "Data cover-up investigation"
title: "Disappearance of data related to the 'Mysterious spike' of deaths after vaccination among people under 45 years of age in New Brunswick"
subtitle: "Investigation report"
output:
  html_document: default
  self_contained: yes
  mode: selfcontained
  always_allow_html: true
  pdf_document: default
#   fig_caption: yes
# toc: no
# toc_depth: 2
# toc_float: no
# number_sections: false
date: '2022-05-29'
author: Generated by [IVIM](https://IVIM.ca)
---

<!-- ## Generated by [IVIM](https://IVIM.ca) -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
<!-- ## Summary -->

<!-- It  -->



## Background

See the following articles and YouTube recording for the background:

- CBC article: "Mysterious spike in N.B. deaths in 2021 not due to COVID, health minister says", Posted: May 13, 2022, URL: https://www.cbc.ca/amp/1.6451680

- «"Mysterious spike" in excess deaths in New Brunswick: Now you see it. Now you don't»,   May 21 <https://ivim.substack.com/p/mysterious-spike-in-excess-deaths>
- «"Killer on the road" in New Brunswick: killing mainly under 44. CBC started talking about it. StatsCan removed data », 
May 19:  <https://ivim.substack.com/p/killer-on-the-road-in-new-brunswick>

- Recording of the "Open Canada Data" seminars on 2022-05-20 : <https://youtu.be/Q3J_P5yk_kM>

## Data source 

Source: Statistics Canada - Table 13-10-0768-01

URL: <https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310076801>  
Data Releases: 2022-04-14 vs. 2022-05-12    
Cached: <https://github.com/IVI-M/std-data/tree/main/ca/statcan-deaths-by-age/NB>




## Data preview

Extracts below show :

- the first and last five lines of the dataset after 2018-01-01, 
- the date of the last entry, 
- the date of the last numeric entry for each age group (i.e. which is when entry is a  number, not NA - Not Available)



### Data Release: 2022-04-14



```{r april, echo=FALSE, message=FALSE, warning=FALSE}
source("00_common.R")
# > Read Deaths By age ----

# https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310076801

if (F) {
  dt <- cansim::get_cansim("13-10-0768-01") %>%  setDT() 
  
    fwrite(dt, "data/13100768-2022-06-09.csv")
  # fwrite(dt, "data/13100768-since2018-2022-05-12.csv")
  saveRDS(dt, paste0("3-10-0768-01-cached_on-2022-May-19.Rds")) # save locally as compressed Rds file
  dt <- dt[, c("Date", "GEO", "val_norm","Age at time of death", "Sex", "Characteristics")] 
}
if (F) {
  dt <- fread("data/13100768-2022-04-14.csv")
  dt <- fread("data/13100768-2022-05-12.csv")
  dt %>% names
  dt <- dt[, c("REF_DATE", "GEO", "VALUE","Age at time of death", "Sex")] 
  setnames(dt, "VALUE", "value")
  setnames(dt, "REF_DATE", "Date")
  setnames(dt, "Age at time of death", "Age")
  dt[, GEO := gsub(", place of occurrence", "", GEO)]
  dt[, Age := gsub("Age at time of death, ", "", Age)]
  dt <- dt[Date>=ymd("2018-01-01")]
  # dt <- dt[Sex=="Both sexes"]
  
  fwrite(dt, "data13100768-since2018-2022-05-12-all-sexes.csv")
  dt <- dt[GEO=="New Brunswick"]
  fwrite(dt, "13100768-NB-since2018-2022-04-14-all-sexes.csv")
  fwrite(dt, "13100768-NB-since2018-2022-05-12-all-sexes.csv")
  fwrite(dt, "13100768-NB-since2018-2022-06-09-all-sexes.csv")
}

dt <- fread("13100768-NB-since2018-2022-04-14-all-sexes.csv")

dt %>% datatable.title ("Total Deaths in New Brunswick before and after vaccination. Data release: 2022-04-14")


dt 

# Release date: 2022-05-12
# Loaded On "2022-05-19"
dt [, max(Date, na.rm = T), by=GEO] #2022-03-05
dt [!is.na(value) & GEO=="New Brunswick", max(Date, na.rm = T), by=.(Age,Sex)]
# 4:     0 to 44 years Both sexes 2021-06-05
# 5:     0 to 44 years      Males 2021-10-09
# 6:     0 to 44 years    Females 2021-11-13
# 7:    45 to 64 years Both sexes 2021-11-20

# Release date: 2022-04-14 . 2022-02-05
# 4:     0 to 44 years Both sexes 2021-10-23
# 5:     0 to 44 years      Males 2021-10-23
# 6:     0 to 44 years    Females 2021-10-23
# 7:    45 to 64 years Both sexes 2021-10-23



dtHistorical <- dt [Date >= ymd("2018-01-01") & Date <= ymd("2019-12-31") ][ , week:=week(Date)
][ ,.(ave=as.integer(mean(value, na.rm=T))), by=.(GEO, Age, week, Sex) ]

dtPandemic <- dt [Date >= ymd("2020-01-01") ][, week:=week(Date)]

dtExcess <- dtHistorical[dtPandemic, on=c( "GEO" , "week",  "Age", "Sex" ) ]

dtExcess[, deaths := frollmean(value, 6, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ]
dtExcess[, ave := frollmean(ave, 8, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ] 

g1 <- ggplot(dtExcess) + 
  geom_vline(xintercept = ymd("2020-12-14"), linetype=1, alpha=0.4) +
  geom_vline(xintercept = ymd("2021-05-01"), linetype=1, size=2, alpha=0.4) +
  # geom_line(aes(Date, deathsMonthly/30*7), data=dt1[Date> "2020-01-01"], col="red",  alpha=0.8) +
  geom_line(aes(Date, value), col="red", alpha=0.3) +
  geom_line(aes(Date, deaths), col="red") +
  geom_line(aes(Date, ave), col="black", alpha=0.5) +
  
  
  facet_grid(Age~Sex, scales = "free") +
  labs( title = "Mortality rate before and after vaccination",
        subtitle="Region: New Brunswick. Data Release: 2022-04-14",
        # subtitle="Data Releases: 2022-04-14",
        
        x = "Black line: pre-pandemic 2018-2019 average. Thick line: six-week average",
        # y = ifelse(F,  "Deaths per million per week", "Deaths per week"),
        y="Deaths per week", # 13-10-0768-01 (age); 13-10-0810-01 (cause)
        caption = "Source: Statistics Canada - Table 13-10-0768-01\n Generated by Open Canada Deaths Tracker App (ivim.ca/app)"     )  

# ggsave(paste0("NB-excess-deaths_asof_2022-05-12.png"), width = 12, height = 9)


```



### Data Release: 2022-05-12



```{r may, echo=FALSE, message=FALSE, warning=FALSE}

# > Read Deaths By age ----


dt <- fread("13100768-NB-since2018-2022-05-12-all-sexes.csv")
dt 

# Release date: 2022-05-12
# Loaded On "2022-05-19"
dt [, max(Date, na.rm = T), by=GEO] #2022-03-05
dt [!is.na(value) & GEO=="New Brunswick", max(Date, na.rm = T), by=.(Age,Sex)]
# 4:     0 to 44 years Both sexes 2021-06-05
# 5:     0 to 44 years      Males 2021-10-09
# 6:     0 to 44 years    Females 2021-11-13
# 7:    45 to 64 years Both sexes 2021-11-20

# Release date: 2022-04-14 . 2022-02-05
# 4:     0 to 44 years Both sexes 2021-10-23
# 5:     0 to 44 years      Males 2021-10-23
# 6:     0 to 44 years    Females 2021-10-23
# 7:    45 to 64 years Both sexes 2021-10-23



dtHistorical <- dt [Date >= ymd("2018-01-01") & Date <= ymd("2019-12-31") ][ , week:=week(Date)
][ ,.(ave=as.integer(mean(value, na.rm=T))), by=.(GEO, Age, week, Sex) ]

dtPandemic <- dt [Date >= ymd("2020-01-01") ][, week:=week(Date)]

dtExcess <- dtHistorical[dtPandemic, on=c( "GEO" , "week",  "Age", "Sex" ) ]

dtExcess[, deaths := frollmean(value, 6, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ]
dtExcess[, ave := frollmean(ave, 8, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ] 

g <- ggplot(dtExcess) + 
  geom_vline(xintercept = ymd("2020-12-14"), linetype=1, alpha=0.4) +
  geom_vline(xintercept = ymd("2021-05-01"), linetype=1, size=2, alpha=0.4) +
  # geom_line(aes(Date, deathsMonthly/30*7), data=dt1[Date> "2020-01-01"], col="red",  alpha=0.8) +
  geom_line(aes(Date, value), col="red", alpha=0.3) +
  geom_line(aes(Date, deaths), col="red") +
  geom_line(aes(Date, ave), col="black", alpha=0.5) +
  
  
  facet_grid(Age~Sex, scales = "free") +
  labs( title = "Mortality rate before and after vaccination",
        subtitle="Region: New Brunswick. Data Release: 2022-05-12",
        # subtitle="Data Releases: 2022-04-14",
        
        x = "Black line: pre-pandemic 2018-2019 average. Thick line: six-week average",
        # y = ifelse(F,  "Deaths per million per week", "Deaths per week"),
        y="Deaths per week", # 13-10-0768-01 (age); 13-10-0810-01 (cause)
        caption = "Source: Statistics Canada - Table 13-10-0768-01\n Generated by Open Canada Deaths Tracker App (ivim.ca/app)"     )  
# g
# ggsave(paste0("NB-excess-deaths_asof_2022-05-12.png"), width = 12, height = 9)


```

## Comparative graphs


<!-- ### Plots for printing -->

Vertical lines indicate the date when vaccination started with high-risk population (December 14, 2020) and (thick line) when vaccination was extended to general population  (May 1, 2021)


```{r comparison, echo=FALSE, message=FALSE, warning=FALSE}

g1
g

```


<!-- ### Interactive plots -->

<!-- Hover over the image for details and selection -->

<!-- ```{r  echo=FALSE, message=FALSE, warning=FALSE} -->
<!-- g1 %>% ggplotly -->
<!-- g %>% ggplotly -->
<!-- ``` -->

## Observations:

- In the release of 2022-05-12 the data for  'all sexes' in age group 0 to 44 after 6 June 2021 have been erased, i.e.. original values that were published on 2022-04-14  replaced with blanks ('NA' (Not available).
Incidentally, this is exactly when  the highest peak in deaths in this population of new-brunswicker was observed and commented on by the province health minister



### Related consideration

The data in several fields of the data are interrelated. 
One can not easily erase data in one field without affecting values in other field. Specifically:

- 'All ages' values should equal to the sum of deaths in all age groups. In April release of the data, this appears to be the case. In May release it is not.

- 'Both sexes' values should be equal to the sum of deaths in both sexes. This does not appear to be the case. It is not clear how or why this could have been possible. 

Hence further investigation and monitoring of these data is recommended

<!-- #### Open questions -->

<!-- It is  -->

<!-- #### Next steps -->