---
title: "New Brunswick excess deaths cover-up investigation"
output:
  pdf_document: default
  html_document: default
  always_allow_html: true
date: '2022-05-29'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

See the following articles and youtube recording for the background:

- CBC article: "Mysterious spike in N.B. deaths in 2021 not due to COVID, health minister says", Posted: May 13, 2022, URL: https://www.cbc.ca/amp/1.6451680

- «"Mysterious spike" in excess deaths in New Brunswick: Now you see it. Now you don't»,   May 21 <https://ivim.substack.com/p/mysterious-spike-in-excess-deaths>
- «"Killer on the road" in New Brunswick: killing mainly under 44. CBC started talking about it. StatsCan removed data », 
May 19:  <https://ivim.substack.com/p/killer-on-the-road-in-new-brunswick>

- Recording of the "Open Canada Data" seminars on 2022-05-20 : <https://youtu.be/Q3J_P5yk_kM>

## Data source 

Source: Statistics Canada - Table 13-10-0768-01

URL: <https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310076801>





## Data preview

Extracts below show :
- the first and last five lines of the dataset after 2018-01-01, 
- the date of the last entry, 
- the date of the last numeric entry for each age group (i.e. which is when entry is a  number, not NA - Not Available)

### Data Release: 2022-04-14



```{r april, echo=FALSE, message=FALSE, warning=FALSE}
source("00_common.R")
# > Read Deaths By age ----

# https://www150.statcan.gc.ca/t1/tbl1/en/tv.action?pid=1310076801

if (F) {
  dt <- cansim::get_cansim("13-10-0768-01") %>%  setDT() 
  fwrite(dt, "data/13100768-since2018-2022-05-12.csv")
  saveRDS(dt, paste0("3-10-0768-01-cached_on-2022-May-19.Rds")) # save locally as compressed Rds file
  dt <- dt[, c("Date", "GEO", "val_norm","Age at time of death", "Sex", "Characteristics")] 
}
if (F) {
  dt <- fread("data/13100768-2022-04-14.csv")
  dt <- fread("data/13100768-2022-05-12.csv")
  dt %>% names
  dt <- dt[, c("REF_DATE", "GEO", "VALUE","Age at time of death", "Sex")] 
  setnames(dt, "VALUE", "value")
  setnames(dt, "REF_DATE", "Date")
  setnames(dt, "Age at time of death", "Age")
  dt[, GEO := gsub(", place of occurrence", "", GEO)]
  dt[, Age := gsub("Age at time of death, ", "", Age)]
  dt <- dt[Date>=ymd("2018-01-01")]
  # dt <- dt[Sex=="Both sexes"]
  fwrite(dt, "13100768-since2018-2022-05-12-all-sexes.csv")
  dt <- dt[GEO=="New Brunswick"]
  fwrite(dt, "13100768-NB-since2018-2022-04-14-all-sexes.csv")
    fwrite(dt, "13100768-NB-since2018-2022-05-12-all-sexes.csv")
}

dt <- fread("13100768-NB-since2018-2022-04-14-all-sexes.csv")
dt 

# Release date: 2022-05-12
# Loaded On "2022-05-19"
dt [, max(Date, na.rm = T), by=GEO] #2022-03-05
dt [!is.na(value) & GEO=="New Brunswick", max(Date, na.rm = T), by=.(Age,Sex)]
# 4:     0 to 44 years Both sexes 2021-06-05
# 5:     0 to 44 years      Males 2021-10-09
# 6:     0 to 44 years    Females 2021-11-13
# 7:    45 to 64 years Both sexes 2021-11-20

# Release date: 2022-04-14 . 2022-02-05
# 4:     0 to 44 years Both sexes 2021-10-23
# 5:     0 to 44 years      Males 2021-10-23
# 6:     0 to 44 years    Females 2021-10-23
# 7:    45 to 64 years Both sexes 2021-10-23



dtHistorical <- dt [Date >= ymd("2018-01-01") & Date <= ymd("2019-12-31") ][ , week:=week(Date)
][ ,.(ave=as.integer(mean(value, na.rm=T))), by=.(GEO, Age, week, Sex) ]

dtPandemic <- dt [Date >= ymd("2020-01-01") ][, week:=week(Date)]

dtExcess <- dtHistorical[dtPandemic, on=c( "GEO" , "week",  "Age", "Sex" ) ]

dtExcess[, deaths := frollmean(value, 6, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ]
dtExcess[, ave := frollmean(ave, 8, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ] 

g1 <- ggplot(dtExcess) + 
  geom_vline(xintercept = ymd("2020-12-14"), linetype=1, alpha=0.4) +
  geom_vline(xintercept = ymd("2021-05-01"), linetype=1, size=2, alpha=0.4) +
  # geom_line(aes(Date, deathsMonthly/30*7), data=dt1[Date> "2020-01-01"], col="red",  alpha=0.8) +
  geom_line(aes(Date, value), col="red", alpha=0.3) +
  geom_line(aes(Date, deaths), col="red") +
  geom_line(aes(Date, ave), col="black", alpha=0.5) +
  
  
  facet_grid(Age~Sex, scales = "free") +
  labs( title = "Mortality rate before and after vaccination",
        subtitle="Region: New Brunswick. Data Release: 2022-04-14",
        # subtitle="Data Releases: 2022-04-14",
        
        x = "Black line: pre-pandemic 2018-2019 average. Thick line: six-week average",
        # y = ifelse(F,  "Deaths per million per week", "Deaths per week"),
        y="Deaths per week", # 13-10-0768-01 (age); 13-10-0810-01 (cause)
        caption = "Source: Statistics Canada - Table 13-10-0768-01\n Generated by Open Canada Deaths Tracker App (ivim.ca/app)"     )  

# ggsave(paste0("NB-excess-deaths_asof_2022-05-12.png"), width = 12, height = 9)


```



### Data Release: 2022-05-12



```{r may, echo=FALSE, message=FALSE, warning=FALSE}

# > Read Deaths By age ----


dt <- fread("13100768-NB-since2018-2022-05-12-all-sexes.csv")
dt 

# Release date: 2022-05-12
# Loaded On "2022-05-19"
dt [, max(Date, na.rm = T), by=GEO] #2022-03-05
dt [!is.na(value) & GEO=="New Brunswick", max(Date, na.rm = T), by=.(Age,Sex)]
# 4:     0 to 44 years Both sexes 2021-06-05
# 5:     0 to 44 years      Males 2021-10-09
# 6:     0 to 44 years    Females 2021-11-13
# 7:    45 to 64 years Both sexes 2021-11-20

# Release date: 2022-04-14 . 2022-02-05
# 4:     0 to 44 years Both sexes 2021-10-23
# 5:     0 to 44 years      Males 2021-10-23
# 6:     0 to 44 years    Females 2021-10-23
# 7:    45 to 64 years Both sexes 2021-10-23



dtHistorical <- dt [Date >= ymd("2018-01-01") & Date <= ymd("2019-12-31") ][ , week:=week(Date)
][ ,.(ave=as.integer(mean(value, na.rm=T))), by=.(GEO, Age, week, Sex) ]

dtPandemic <- dt [Date >= ymd("2020-01-01") ][, week:=week(Date)]

dtExcess <- dtHistorical[dtPandemic, on=c( "GEO" , "week",  "Age", "Sex" ) ]

dtExcess[, deaths := frollmean(value, 6, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ]
dtExcess[, ave := frollmean(ave, 8, align = "center", fill = NA, na.rm=T),by=.(Age,Sex) ] 

g <- ggplot(dtExcess) + 
  geom_vline(xintercept = ymd("2020-12-14"), linetype=1, alpha=0.4) +
  geom_vline(xintercept = ymd("2021-05-01"), linetype=1, size=2, alpha=0.4) +
  # geom_line(aes(Date, deathsMonthly/30*7), data=dt1[Date> "2020-01-01"], col="red",  alpha=0.8) +
  geom_line(aes(Date, value), col="red", alpha=0.3) +
  geom_line(aes(Date, deaths), col="red") +
  geom_line(aes(Date, ave), col="black", alpha=0.5) +
  
  
  facet_grid(Age~Sex, scales = "free") +
  labs( title = "Mortality rate before and after vaccination",
        subtitle="Region: New Brunswick. Data Release: 2022-04-14",
        # subtitle="Data Releases: 2022-04-14",
        
        x = "Black line: pre-pandemic 2018-2019 average. Thick line: six-week average",
        # y = ifelse(F,  "Deaths per million per week", "Deaths per week"),
        y="Deaths per week", # 13-10-0768-01 (age); 13-10-0810-01 (cause)
        caption = "Source: Statistics Canada - Table 13-10-0768-01\n Generated by Open Canada Deaths Tracker App (ivim.ca/app)"     )  
# g
# ggsave(paste0("NB-excess-deaths_asof_2022-05-12.png"), width = 12, height = 9)


```

## Comparative graphs


### Plots for printing

Vertical lines indicate tghe date when vaccination started with high-risk population (December 14, 2020) and further extended to general population (May 1, 2021)


```{r comparison, echo=FALSE, message=FALSE, warning=FALSE}

g1
g

```


<!-- ### Interactive plots -->

<!-- Hover over the image for details and selection -->

<!-- ```{r  echo=FALSE, message=FALSE, warning=FALSE} -->
<!-- g1 %>% ggplotly -->
<!-- g %>% ggplotly -->
<!-- ``` -->

## Observations:

- In May data release on May 12, 2022,  'all sexes' data in age group 0 to 44 have been erased (values replaced with 'NA') after June 6, 2021, which is when the highest peak in deaths was previously registered.



### Related consideration and findings

The data in the fields (groups) are interelated. 
One can not easily erase data in one group without affecting values in other groups.

- 'All ages' values should equal to the sum of deaths in all age groups. In April release of the data, this apepars to be the case. In May release it is not.

- 'Both sexes' values should equal to the sum of deaths in both sexes. This does not appear to the case. This could be a result of non-intentional human arro

<!-- Open questions -->

It is 